---
title: "HUMAnN3 Functional Profiling with Differential Analysis"
author: "Luicer Olubayo"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document:
    toc: true
    toc_depth: 3
---

## Overview

This practical introduces students to HUMAnN3-derived functional microbiome outputs using pathway abundance, pathway coverage, and gene family tables.

### Questions addressed in this session

1. What microbial functions are most abundant across the cohort?
2. Do samples cluster by clinical group based on functional profiles?
3. Which broad biological categories dominate the microbiome?
4. Which pathways differ across groups?
5. Which gene families differ across groups?
6. Do the results highlight biology linked to the paper, such as metabolism, respiration, iron/heme handling, and nutrient utilization?

### Expected learning points

- HUMAnN outputs include both broad functional summaries and high-resolution gene family features.
- Pathway abundance and pathway coverage answer different questions.
- Gene-family analysis requires filtering because of the very large number of features.
- Exploratory plots show broad structure, while differential abundance analysis identifies specific group-associated functions.

## Prerequisites

This module assumes:

- Basic familiarity with R and RStudio
- Completed upstream processing (e.g., BWA, Prokka, Sylph, MetaPhlAn, HUMAnN3)
- Availability of:
  - pathway abundance tables
  - pathway coverage tables
  - gene family tables
  - sample metadata

Required R packages:
- tidyverse
- pheatmap
- vegan
- Maaslin2

## Input data structure

- Rows = features (pathways or gene families)
- Columns = samples
- First column = Feature IDs
- Metadata must include:
  - sample_name
  - run_accession
  - group

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 6,
  dpi = 300
)
```


```{r libraries}
# ============================================
# LIBRARIES
# ============================================

library(tidyverse)
library(pheatmap)
library(vegan)

if (!requireNamespace("Maaslin2", quietly = TRUE)) {
  if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
  }
  BiocManager::install("Maaslin2")
}
library(Maaslin2)
```

```{r helper-functions}
# ============================================
# HELPER-FUNCTIONS
# ============================================

extract_run_id <- function(x) {
  x <- gsub("_Abundance-RPKs$", "", x)
  x <- gsub("_Abundance$", "", x)
  x <- gsub("_Coverage$", "", x)
  x
}

clean_names <- function(x) {
  x <- gsub("^.*?:\\s*", "", x)   # remove pathway ID prefix
  x <- gsub("\\(.*?\\)", "", x)   # remove bracketed text like (E. coli)
  x <- gsub("_", " ", x)
  trimws(x)
}

annotate_pathways <- function(x) {
  case_when(
    grepl("glycolysis|fermentation|respiration|TCA|energy|formate", x, ignore.case = TRUE) ~ "Energy metabolism",
    grepl("amino acid|valine|leucine|isoleucine|ornithine|biosynthesis", x, ignore.case = TRUE) ~ "Amino acid metabolism",
    grepl("nucleotide|purine|pyrimidine|UMP|adenine|guanosine|allantoin", x, ignore.case = TRUE) ~ "Nucleotide metabolism",
    grepl("peptidoglycan|cell wall|muramoyl|lipopolysaccharide", x, ignore.case = TRUE) ~ "Cell wall",
    grepl("carbohydrate|glycogen|sugar|fucose|galactarate", x, ignore.case = TRUE) ~ "Carbohydrate metabolism",
    grepl("cofactor|vitamin|heme|iron|aerobactin|ubiquinol|menaquinol", x, ignore.case = TRUE) ~ "Cofactors, iron & respiration",
    TRUE ~ "Other"
  )
}

remove_zero_variance_features <- function(mat) {
  keep <- apply(mat, 1, function(x) {
    x <- x[!is.na(x)]
    length(x) > 1 && var(x) > 0
  })
  mat[keep, , drop = FALSE]
}
```

```{r load-data}
# ============================================
# LOAD DATA
# ============================================

path_abun <- read.delim(
  "/Users/bg7/Library/CloudStorage/GoogleDrive-bg7@sanger.ac.uk/My Drive/CHAIN Culture Collection/r_scripts/humann3_pathabundance_merged_normalised_UG.tsv",
  sep = "\t",
  check.names = FALSE
)

path_cov <- read.delim(
  "/Users/bg7/Library/CloudStorage/GoogleDrive-bg7@sanger.ac.uk/My Drive/CHAIN Culture Collection/r_scripts/humann3_pathcoverage_merged_UG.tsv",
  sep = "\t",
  check.names = FALSE
)

genefam <- read.delim(
  "/Users/bg7/Library/CloudStorage/GoogleDrive-bg7@sanger.ac.uk/My Drive/CHAIN Culture Collection/r_scripts/humann3_genefamilies_merged_normalised_UG.tsv",
  sep = "\t",
  check.names = FALSE
)

meta <- read.csv(
  "/Users/bg7/Library/CloudStorage/GoogleDrive-bg7@sanger.ac.uk/My Drive/CHAIN Culture Collection/r_scripts/Bednarski_2025_cleaned_metadata_final.csv",
  check.names = FALSE
)

dim(path_abun)
dim(path_cov)
dim(genefam)
dim(meta)
head(meta)
```

```{r clean-data}
# ============================================
# CLEAN DATA
# ============================================

colnames(path_abun)[1] <- "Feature"
colnames(path_cov)[1]  <- "Feature"
colnames(genefam)[1]   <- "Feature"

path_abun <- path_abun %>%
  filter(!grepl("UNMAPPED|UNINTEGRATED", Feature))

path_cov <- path_cov %>%
  filter(!grepl("UNMAPPED|UNINTEGRATED", Feature))

genefam <- genefam %>%
  filter(!grepl("UNMAPPED|UNINTEGRATED", Feature))

# Keep unstratified rows for main functional analysis
path_abun_unstrat <- path_abun %>%
  filter(!grepl("\\|", Feature))

path_cov_unstrat <- path_cov %>%
  filter(!grepl("\\|", Feature))
```

```{r sample-name-check}
# ============================================
# CHECK COLUMN NAMING
# ============================================

head(colnames(path_abun_unstrat))
head(colnames(path_cov_unstrat))
head(colnames(genefam))

ncol(path_abun_unstrat)
ncol(path_cov_unstrat)
ncol(genefam)
```

```{r match-metadata}
# ============================================
# MATCH METADATA TO HUMAnN SAMPLES
# ============================================

meta2 <- meta %>%
  filter(!is.na(metagm_id), metagm_id != "")

abun_ids <- extract_run_id(colnames(path_abun_unstrat)[-1])
cov_ids  <- extract_run_id(colnames(path_cov_unstrat)[-1])
gene_ids <- extract_run_id(colnames(genefam)[-1])

cat("Pathway abundance samples:", length(abun_ids), "\n")
cat("Pathway coverage samples:", length(cov_ids), "\n")
cat("Gene family samples:", length(gene_ids), "\n")
cat("Metadata samples with metagm_id:", nrow(meta2), "\n")

shared_ids <- Reduce(intersect, list(abun_ids, cov_ids, gene_ids, meta2$metagm_id))
shared_ids <- abun_ids[abun_ids %in% shared_ids]

cat("Shared samples across all data types and metadata:", length(shared_ids), "\n")

abun_keep <- match(shared_ids, abun_ids)
cov_keep  <- match(shared_ids, cov_ids)
gene_keep <- match(shared_ids, gene_ids)

path_abun_unstrat <- path_abun_unstrat[, c(1, abun_keep + 1), drop = FALSE]
path_cov_unstrat  <- path_cov_unstrat[, c(1, cov_keep + 1), drop = FALSE]
genefam <- genefam[, c(1, gene_keep + 1), drop = FALSE]

meta2 <- meta2[match(shared_ids, meta2$metagm_id), , drop = FALSE]

stopifnot(all(extract_run_id(colnames(path_abun_unstrat)[-1]) == shared_ids))
stopifnot(all(extract_run_id(colnames(path_cov_unstrat)[-1]) == shared_ids))
stopifnot(all(extract_run_id(colnames(genefam)[-1]) == shared_ids))
stopifnot(all(meta2$run_accession == shared_ids))

meta <- meta2
meta$sick <- as.factor(meta$sick)

colnames(path_abun_unstrat)[-1] <- meta$sample_name
colnames(path_cov_unstrat)[-1]  <- meta$sample_name
colnames(genefam)[-1]           <- meta$sample_name

cat("Final matched sample count:", nrow(meta), "\n")
table(meta$sick)
head(meta[, c("sample_name", "metagm_id", "sick", "country")])
```
## Question 1: What are the most abundant microbial functions across the cohort?

These first plots are exploratory. They help us understand broad functional structure before testing specific associations.
```{r pathway-matrix}
# ============================================
# PATHWAY MATRIX
# ============================================

path_mat <- path_abun_unstrat %>%
  column_to_rownames("Feature") %>%
  as.matrix()

mode(path_mat) <- "numeric"
path_mat_log <- log10(path_mat + 1e-6)
```

We begin by examining global structure using dimensionality reduction.
Principal Component Analysis (PCA) allows us to assess whether samples cluster by group.

```{r pathway-pca}
# ============================================
# PCA OF PATHWAY ABUNDANCE
# ============================================

pca_input <- t(path_mat_log)

keep_cols <- apply(
  pca_input,
  2,
  function(x) {
    x <- x[!is.na(x)]
    length(x) > 1 && var(x) > 0
  }
)

pca_input <- pca_input[, keep_cols, drop = FALSE]

cat("Samples used for PCA:", nrow(pca_input), "\n")
cat("Pathways retained for PCA:", ncol(pca_input), "\n")

pca <- prcomp(pca_input, scale. = TRUE)

meta <- meta[match(rownames(pca$x), meta$sample_name), ]
meta$sick <- factor(meta$sick, levels = c("SM", "CC"))

{group_colors <- c("SM" = "#00BFC4", "CC" = "#F8766D")
point_colors <- group_colors[meta$sick]

plot(
  pca$x[, 1], pca$x[, 2],
  col = point_colors,
  pch = 19,
  xlab = paste0("PC1 (", round(100 * summary(pca)$importance[2, 1], 1), "%)"),
  ylab = paste0("PC2 (", round(100 * summary(pca)$importance[2, 2], 1), "%)"),
  main = "PCA of Pathway Abundance"
)

legend(
  "topright",
  legend = names(group_colors),
  col = group_colors,
  pch = 19,
  title = "Group"
)}

table(meta$sick)

```

**Interpretation prompt:**  
Do samples separate clearly by group, or is the clustering partial?
If separation is incomplete, that suggests the functional shifts are subtle rather than absolute.

```{r top-pathway-heatmap}
# ============================================
# TOP ABUNDANT PATHWAYS
# ============================================

path_means <- rowMeans(path_mat, na.rm = TRUE)
top_pathways <- names(sort(path_means, decreasing = TRUE))[1:30]

heat_path <- path_mat_log[top_pathways, , drop = FALSE]
heat_path_scaled <- t(scale(t(heat_path)))

rownames(heat_path_scaled) <- clean_names(rownames(heat_path_scaled))

# Optional: colors for annotation
ann_colors <- list(
  Sick = c(
    SM = "#00BFC4",
    CC    = "#F8766D"
  )
)

pheatmap(
  heat_path_scaled,
  show_rownames = TRUE,
  show_colnames = FALSE,
  main = "Top 30 Most Abundant Unstratified Pathways (Scaled)",
  annotation_col = annotation_col,
  annotation_colors = ann_colors
)

```

**Interpretation prompt:**  
These pathways usually reflect core metabolism and microbial growth rather than group-specific biology.

## Question 2: How consistently are pathways detected across samples?

```{r pathway-coverage}
# ============================================
# PATHWAY COVERAGE
# ============================================

cov_mat <- path_cov_unstrat %>%
  column_to_rownames("Feature") %>%
  as.matrix()

mode(cov_mat) <- "numeric"

prev <- rowSums(cov_mat > 0, na.rm = TRUE)
cov_filtered <- cov_mat[prev > 0.2 * ncol(cov_mat), , drop = FALSE]

coverage_summary <- data.frame(
  Feature = rownames(cov_mat),
  Prevalence = prev,
  Proportion = prev / ncol(cov_mat)
) %>%
  arrange(desc(Proportion))

dim(cov_filtered)
head(coverage_summary, 20)
```

**Interpretation prompt:**  
Pathway abundance asks how much of a pathway is present.
Pathway coverage asks how consistently a pathway is represented across samples.

## Question 3: Which broad biological systems dominate the microbiome?

```{r functional-categories}
# ============================================
# FUNCTIONAL CATEGORIES
# ============================================

pathway_names <- clean_names(rownames(path_mat))
categories <- annotate_pathways(pathway_names)

table(categories)

cat_df <- data.frame(category = categories)

ggplot(cat_df, aes(x = category)) +
  geom_bar(fill = "steelblue") +
  theme_bw() +
  coord_flip() +
  labs(
    title = "Distribution of Functional Pathway Categories",
    x = "Category",
    y = "Number of pathways"
  )
```

**Interpretation prompt:**  
This plot shows how many pathways fall into each broad biological category.
It helps move from pathway IDs to interpretable biology.

```{r category-mean-heatmap}
# ============================================
# MEAN ABUNDANCE BY FUNCTIONAL CATEGORY
# ============================================

path_df <- as.data.frame(path_mat)
path_df$category <- categories

category_abundance_mean <- path_df %>%
  group_by(category) %>%
  summarise(across(where(is.numeric), mean))

cat_mat_mean <- category_abundance_mean %>%
  column_to_rownames("category") %>%
  as.matrix()

cat_mat_mean_log <- log10(cat_mat_mean + 1e-6)

pheatmap(
  cat_mat_mean_log,
  main = "Mean Pathway Abundance by Functional Category",
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  show_colnames = FALSE,
  annotation_col = annotation_col,
  annotation_colors = ann_colors
)
```

```{r category-mean-heatmap-no-other}
# ============================================
# EXCLUDING THE BROAD 'OTHER' CATEGORY
# ============================================

cat_mat_mean_no_other <- cat_mat_mean[rownames(cat_mat_mean) != "Other", , drop = FALSE]

pheatmap(
  log10(cat_mat_mean_no_other + 1e-6),
  main = "Mean Pathway Abundance by Functional Category (Excluding Other)",
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  show_colnames = FALSE,
  annotation_col = annotation_col,
  annotation_colors = ann_colors
)
```

**Interpretation prompt:**  
This aggregation is more interpretable than individual features and makes it easier to discuss pathways in terms of metabolism, cell wall processes, and nucleotide or energy functions.

## Question 4: Do broad biological categories differ across groups?

```{r category-boxplot}
# ============================================
# CATEGORY ABUNDANCE BY GROUP
# ============================================

cat_df_group <- as.data.frame(t(cat_mat_mean))
cat_df_group$sick <- meta$sick

cat_long <- cat_df_group %>%
  pivot_longer(-sick, names_to = "category", values_to = "abundance")

ggplot(cat_long, aes(x = sick, y = log10(abundance + 1e-6), fill = sick)) +
  geom_boxplot() +
  facet_wrap(~category, scales = "free_y") +
  theme_bw() +
  labs(
    title = "Functional Categories by Group",
    x = "Group",
    y = "log10(mean category abundance + 1e-6)"
  )
```

**Interpretation prompt:**  
Do certain groups show relatively higher energy metabolism, nucleotide metabolism, or cell-wall-associated pathways?
These plots summarize broad biology before moving into feature-level testing.

## Question 5: Do we see paper-relevant functional themes?

The paper highlights metabolism, respiration, iron/heme handling, and nutrient utilization. We can inspect those themes directly.

```{r paper-focused-pathways}
# ============================================
# PAPER-FOCUSED PATHWAYS
# ============================================

clean_features <- clean_names(rownames(path_mat))

target_idx <- grepl(
  "respiration|iron|heme|glycolysis|fermentation|aerobactin|fucose|sialic|allantoin|ethanolamine|ubiquinol|menaquinol",
  clean_features,
  ignore.case = TRUE
)

target_mat <- path_mat[target_idx, , drop = FALSE]
rownames(target_mat) <- clean_names(rownames(target_mat))
target_log <- log10(target_mat + 1e-6)

# Create annotation for columns (samples)
annotation_col <- data.frame(
  Sick = meta$sick
)
rownames(annotation_col) <- meta$sample_name  # must match target_log columns

if (nrow(target_log) > 1) {
  pheatmap(
    target_log,
    show_rownames = TRUE,
    show_colnames = FALSE,
    main = "Paper-Relevant Functional Pathways",
    annotation_col = annotation_col,
    annotation_colors = ann_colors
  )
}

```

**Interpretation prompt:**  
Are these functions consistently elevated in some groups?
Do they suggest microbial communities adapted to altered nutrient availability, host-derived substrates, or stressed gut environments?

## Question 6: What happens at the gene-family level?

Gene-family tables are much larger and more granular than pathways, so filtering is essential.

```{r gene-families-filtering}
# ============================================
# GENE FAMILY ANALYSIS
# ============================================

genefam_mat <- genefam %>%
  column_to_rownames("Feature") %>%
  as.matrix()

mode(genefam_mat) <- "numeric"

gene_prev <- rowMeans(genefam_mat > 0, na.rm = TRUE)
gene_mean <- rowMeans(genefam_mat, na.rm = TRUE)

genefam_filt <- genefam_mat[
  gene_prev >= 0.10 & gene_mean >= 1e-05,
  ,
  drop = FALSE
]

cat("Original gene families:", nrow(genefam_mat), "\n")
cat("Retained after prevalence + mean-abundance filtering:", nrow(genefam_filt), "\n")
```


```{r gene-families-top-heatmap}
# ============================================
# TOP VARIABLE FILTERED GENE FAMILIES
# ============================================

var_genes <- apply(genefam_filt, 1, var, na.rm = TRUE)

n_keep <- min(200, length(var_genes))
top_gene_names <- names(sort(var_genes, decreasing = TRUE))[1:n_keep]

genefam_top <- genefam_filt[top_gene_names, , drop = FALSE]
genefam_log <- log10(genefam_top + 1e-6)

pheatmap(
  genefam_log,
  show_rownames = FALSE,
  show_colnames = FALSE,
  main = "Top Variable Gene Families",
  annotation_col = annotation_col,
  annotation_colors = ann_colors
)

```

**Interpretation prompt:**  
Gene families offer higher resolution, but they are harder to interpret biologically without annotation or targeted grouping.

## Question 7: Are total pathway or gene-family loads different across groups?

```{r global-load-boxplots}
# ============================================
# GLOBAL FUNCTIONAL LOAD
# ============================================

meta$total_pathway_abundance <- colSums(path_mat, na.rm = TRUE)
meta$total_gene_family_abundance <- colSums(genefam_mat, na.rm = TRUE)

p1 <- ggplot(meta, aes(x = sick, y = log10(total_pathway_abundance + 1e-6), fill = sick)) +
  geom_boxplot() +
  theme_bw() +
  labs(
    title = "Total Pathway Abundance by Group",
    x = "Group",
    y = "log10(total pathway abundance + 1e-6)"
  )

p2 <- ggplot(meta, aes(x = sick, y = log10(total_gene_family_abundance + 1e-6), fill = sick)) +
  geom_boxplot() +
  theme_bw() +
  labs(
    title = "Total Gene Family Abundance by Group",
    x = "Group",
    y = "log10(total gene-family abundance + 1e-6)"
  )

p1
p2
```

**Interpretation prompt:**  
These plots ask whether groups differ in the overall amount of annotated functional signal, before focusing on individual features.

## Question 8: Which pathways are significantly associated with group?

```{r prepare-maaslin}
# ============================================
# PREPARE INPUTS FOR MaAsLin2
# ============================================

path_mat_maaslin <- remove_zero_variance_features(path_mat)
genefam_top_maaslin <- remove_zero_variance_features(genefam_top)

path_maaslin <- as.data.frame(t(path_mat_maaslin))
gene_maaslin <- as.data.frame(t(genefam_top_maaslin))

stopifnot(all(rownames(path_maaslin) == meta$sample_name))
stopifnot(all(rownames(gene_maaslin) == meta$sample_name))

meta_maaslin <- meta
rownames(meta_maaslin) <- meta_maaslin$sample_name
meta_maaslin$sick <- as.factor(meta_maaslin$sick)

stopifnot(!any(is.na(meta_maaslin$sick)))
stopifnot(!any(duplicated(rownames(meta_maaslin))))

table(meta_maaslin$sick)

```


```{r maaslin-pathways, eval=FALSE}
# ============================================
# MaAsLin3: PATHWAYS
# ============================================
library(maaslin3)
fit_path <- maaslin3(
  input_data = path_maaslin,
  input_metadata = meta_maaslin,
  output = "maaslin_pathways_group",
  fixed_effects = c("sick"),
  normalization = "NONE",
  transform = "LOG"
)

```
**Note:** These models are pre-run. Results are read from the output folders.

```{r maaslin-genefamilies, eval=FALSE}
# ============================================
# MaAsLin3: GENE FAMILIES
# ============================================

fit_gene <- maaslin3(
  input_data = gene_maaslin,
  input_metadata = meta_maaslin,
  output = "maaslin_genefamilies_group",
  fixed_effects = c("sick"),
  normalization = "NONE",
  transform = "LOG"
)
```

```{r pathway-volcanos}
# ============================================
# VOLCANO PLOT: PATHWAYS
# ============================================

res_path <- read.delim("maaslin_pathways_group/all_results.tsv")
res_path$clean_feature <- clean_names(res_path$feature)

ggplot(res_path, aes(x = coef, y = -log10(qval_individual))) +
  geom_point(aes(color = qval_individual < 0.25), alpha = 0.7) +
  geom_hline(yintercept = -log10(0.25), linetype = "dashed", color = "grey40") +
  theme_bw() +
  labs(
    title = "Volcano Plot: Pathways vs Sick",
    x = "Effect size (coefficient)",
    y = "-log10(q-value)"
  )
```

```{r genefamilies-volcano}
# ============================================
# VOLCANO PLOT: GENE FAMILIES
# ============================================

res_gene <- read.delim("maaslin_genefamilies_group/all_results.tsv")
res_gene$clean_feature <- clean_names(res_gene$feature)

ggplot(res_gene, aes(x = coef, y = -log10(qval_individual))) +
  geom_point(aes(color = qval_individual < 0.10), alpha = 0.7) +
  geom_hline(yintercept = -log10(0.10), linetype = "dashed", color = "grey40") +
  theme_bw() +
  labs(
    title = "Volcano Plot: Gene Families vs Sick",
    x = "Effect size (coefficient)",
    y = "-log10(q-value)"
  )
```

## Question 9: Which significant pathway results are easiest to interpret biologically?


```{r significant-pathways-tables}
# ============================================
# SIGNIFICANT PATHWAY RESULTS
# ============================================

res_path <- read.delim("maaslin_pathways_group/all_results.tsv", check.names = FALSE)

# Clean labels for display only
res_path <- res_path %>%
  mutate(
    clean_feature = gsub("\\.+", " ", feature) %>% trimws()
  )

# Summary first (teaching moment)
cat("Min q-value:", min(res_path$qval, na.rm = TRUE), "\n")
cat("q < 0.25:", sum(res_path$qval < 0.25, na.rm = TRUE), "\n")
cat("q < 0.10:", sum(res_path$qval < 0.10, na.rm = TRUE), "\n")
cat("q < 0.05:", sum(res_path$qval < 0.05, na.rm = TRUE), "\n")

# Significant set and annotate biology
sig_path <- res_path %>%
  filter(qval_individual < 0.25) %>%
  arrange(qval_individual) %>%
  mutate(category = annotate_pathways(clean_feature))

# PRINT explicitly
knitr::kable(
  sig_path %>%
    select(clean_feature, category, coef, pval_individual, qval_individual) %>%
    slice_head(n = 20),
  caption = "Top 20 pathways associated with Sick"
)

```



```{r top-significant-pathways-heatmaps}
# ============================================
# HEATMAP OF TOP DIFFERENTIAL PATHWAYS
# ============================================

# Map MaAsLin3 feature names back to original HUMAnN names
feature_map <- data.frame(
  original = rownames(path_mat),
  maaslin = make.names(rownames(path_mat)),
  stringsAsFactors = FALSE
)

top_maaslin <- sig_path %>%
  slice_head(n = 20) %>%
  pull(feature)

top_original <- feature_map$original[
  match(top_maaslin, feature_map$maaslin)
]

top_original <- na.omit(top_original)

cat("Mapped features:", length(top_original), "\n")

if (length(top_original) >= 2) {

  abund_top <- path_mat[top_original, , drop = FALSE]
  rownames(abund_top) <- clean_names(rownames(abund_top))

  anno <- meta %>%
  select(sample_name, sick, country) %>%
  as.data.frame()
rownames(anno) <- anno$sample_name
anno$sample_name <- NULL

  pheatmap(
    log10(abund_top + 1e-6),
    annotation_col = anno,
    scale = "row",
    fontsize_row = 8,
    main = "Top Differential Pathways"
  )

} else {
  cat("Not enough features for heatmap\n")
}

###
feature_map <- data.frame(
  original = rownames(path_mat),
  clean = clean_names(rownames(path_mat)),
  stringsAsFactors = FALSE
)

top_clean <- clean_names(top_maaslin)

top_original <- feature_map$original[
  match(top_clean, feature_map$clean)
]

top_original <- na.omit(top_original)

```

**Interpretation prompt:**  
Do the significant features point to broad metabolic shifts, or to specific processes such as respiration, iron handling, or nutrient scavenging?

## Question 10: Can we collapse gene families into targeted biological modules?

This section turns gene families into interpretable functional modules inspired by the paper.

```{r targeted-gene-module}
# ============================================
# TARGETED GENE-FAMILY MODULES
# ============================================

gene_ids <- rownames(genefam_mat)

bucket <- rep("other", length(gene_ids))
bucket[grepl("aerobactin|siderophore|fhu|heme", gene_ids, ignore.case = TRUE)] <- "iron_acquisition"
bucket[grepl("formate dehydrogenase|fdn|ubiquinone|menaquinone|superoxide dismutase", gene_ids, ignore.case = TRUE)] <- "respiration_oxidative_stress"
bucket[grepl("purine|adenosine deaminase|allantoin", gene_ids, ignore.case = TRUE)] <- "purine_metabolism"
bucket[grepl("ethanolamine|fucose|sialic", gene_ids, ignore.case = TRUE)] <- "host_derived_nutrient_use"

table(bucket)

iron_total <- colSums(genefam_mat[bucket == "iron_acquisition", , drop = FALSE], na.rm = TRUE)
resp_total <- colSums(genefam_mat[bucket == "respiration_oxidative_stress", , drop = FALSE], na.rm = TRUE)
purine_total <- colSums(genefam_mat[bucket == "purine_metabolism", , drop = FALSE], na.rm = TRUE)
hostnut_total <- colSums(genefam_mat[bucket == "host_derived_nutrient_use", , drop = FALSE], na.rm = TRUE)

target_df <- data.frame(
  sample_name = colnames(genefam_mat),
  iron_acquisition = as.numeric(iron_total),
  respiration_oxidative_stress = as.numeric(resp_total),
  purine_metabolism = as.numeric(purine_total),
  host_derived_nutrient_use = as.numeric(hostnut_total)
) %>%
  left_join(meta, by = "sample_name")
```

```{r targetted-gene-module-boxplot}
# ============================================
# TARGETED MODULES BY GROUP
# ============================================

target_long <- target_df %>%
  pivot_longer(
    cols = c(iron_acquisition, respiration_oxidative_stress, purine_metabolism, host_derived_nutrient_use),
    names_to = "module",
    values_to = "abundance"
  )

ggplot(target_long, aes(x = sick, y = log10(abundance + 1e-6), fill = sick)) +
  geom_boxplot() +
  facet_wrap(~module, scales = "free_y") +
  theme_bw() +
  labs(
    title = "Targeted Functional Modules by Group",
    x = "Group",
    y = "log10(module abundance + 1e-6)"
  )

```

**Interpretation prompt:**  
Do these targeted functional modules recapitulate the biology emphasized in the paper?

## Reproducibility

To run this module:

1. Clone the repository
2. Place input files in the `data/` directory
3. Open the `.Rmd` file in RStudio
4. Click **Knit → PDF**

Note: MaAsLin3 outputs are precomputed.


