---
title: "Microbial Community Profiling with Differential Analysis"
author: "Bonface Gichuki"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

## Overview

This practical introduces students to taxonomic microbiome analysis using sequencing-derived abundance tables 
-(e.g., from Sylph, focusing on:
  - Community structure
  - Diversity metrics
  - Differential abundance and prevalence

The workflow uses:
- Phyloseq for data handling and diversity analysis
- MaAsLin3 for statistical modelling

## Questions addressed in this session

1. How does alpha diversity differ between children with severe malaria (SM) and Community Controls (CC)?
2. Do samples cluster by clinical group in beta diversity (ordination space)?
3. Which taxa dominate the microbial community at different taxonomic levels?
4. Are there community-wide shifts associated with disease status?
5. Which bacterial taxa are differentially abundant between SM and CC?
6. Do results highlight biologically relevant taxa (e.g. Enterobacteriaceae species)?

## Expected learning points
- Taxonomic profiles describe who is present, not just function
- Alpha diversity reflects within-sample diversity
- Beta diversity reflects between-sample differences
- Ordination methods (MDS (Multidimensional Scaling) ) reveal clustering patterns
- Differential prevalence and abundance requires careful statistical modelling
- Compositionality must be considered (e.g. Centered Log-Ratio (CLR) transformation)

## Prerequisites

This module assumes:

- Basic familiarity with R / RStudio
- A phyloseq object (ps) already constructed
- Data derived from tools such Sylph
- Sample metadata

```{r}
# ============================================
# LIBRARIES AND WORKING DIRECTORY
# ============================================
library(readr)
library(tidyverse)
library(stringr)
library(tibble)
library(data.table)

# Set working directory
wk_path="/Users/bg7/Library/CloudStorage/GoogleDrive-bg7@sanger.ac.uk/My Drive/CHAIN Culture Collection/r_scripts/"
setwd(wk_path)

# ============================================
# Create a sylph relative abundance matrix
# ============================================
# List all Sylph output files
files <- list.files(path = wk_path, pattern = "\\.sylphmpa$", full.names = TRUE)

# Extract sample ID from filename
get_sample_id <- function(path) {
  fname <- basename(path)
  # first drop extension:
  core <- fname %>% str_remove("_clean_1.fastq.gz.sylphmpa") %>% 
    str_remove("prefix_")
  return(core)
}

# Read and stack all Sylph tables
all_sylph <- purrr::map_dfr(
  files,
  function(f) {
    # try skipping comment/meta lines starting with '#'
    df <- fread(f,skip = 1,data.table = F)
    sample_id <- get_sample_id(f)
    df %>%
      # keep only the columns we need
      select(clade_name,relative_abundance) %>%
      mutate(sample = sample_id)})

# Get abundance matrix
abund_mat <- all_sylph %>%
  pivot_wider(id_cols = clade_name, names_from = sample,
    values_from =relative_abundance,
    values_fill = 0         # missing taxa → 0 abundance
  ) %>% arrange(clade_name)

# Save the matrix
write_tsv(abund_mat, "Bednarski_2025_sylph_clade_relative_abundance_matrix.tsv")

```


```{r}
# ============================================
# Create a phyloseq object
# ============================================
library(phyloseq)
library(ggpubr) 
library(microViz)
library(corncob)
library(ggthemes)
library(tidyverse)

# Load sylph abundance matrix and participant metadata
abundance_data <- read.delim("Bednarski_20025_sylph_clade_relative_abundance_matrix.tsv")
metadata <- read.csv("Bednarski_2025_cleaned_metadata_final.csv") %>%
  column_to_rownames("metagm_id")

# Prepare OTU table (rows are taxonomic ranks and columns are samples)
otu_tab <- abundance_data %>%
  column_to_rownames("clade_name")

# Convert otu table into a matrix
otu_mat <- as.matrix(otu_tab)

# only include species level taxonomic ranks
otu_mat_new <- otu_mat[grepl("\\|s__[^|]+$", rownames(otu_mat)), ]
rownames(otu_mat_new)

# Align samples to metadata
common_samples <- intersect(colnames(otu_mat_new), rownames(metadata))
otu_mat_new <- otu_mat_new[, common_samples]
metadata_sub <- metadata[common_samples, ]

# Sanity check
stopifnot(all(colnames(otu_mat_new) == rownames(metadata_sub)))

# Build taxonomy table
tax_split <- strsplit(rownames(otu_mat_new), "\\|")
max_len <- max(lengths(tax_split))
tax_mat <- do.call(rbind, lapply(tax_split, function(x) {
  c(x, rep(NA, max_len - length(x)))
}))

colnames(tax_mat) <- c("Kingdom","Phylum","Class","Order","Family","Genus","Species","Strain")[1:max_len]
rownames(tax_mat) <- rownames(otu_mat_new)

# Create a phyloseq object (includes otu_table, sample metadata and taxonomy table)
ps <- phyloseq(
  otu_table(otu_mat_new, taxa_are_rows = TRUE),
  sample_data(metadata_sub),
  tax_table(as.matrix(tax_mat)))

# Clean metadata (important)
# remove NA in key variable
ps <- subset_samples(ps, !is.na(sick))

# Ensure the grouping variable is a factor
sample_data(ps)$sick <- factor(sample_data(ps)$sick)


# Inspect contents of the phyloseq object (sanity check)
ps
head(otu_table(ps))
head(sample_data(ps))
head(tax_table(ps))
nsamples(ps)
sample_variables(ps)
sample_sums(ps)[1:10] 

```

```{r}
# ============================================
# Community composition (sample-level summary)
# ============================================
library(remotes)
library(devtools)
library(gghalves)
library(microbiomeutilities)

# Clean Family names (remove "f__" prefix for readability in plots)
tax_table(ps)[, "Family"] <- sub("^f__", "", tax_table(ps)[, "Family"])

# Convert phyloseq object to long format for manipulation
df <- psmelt(ps)

# Clean Family names again in melted data (ensures consistency)
df$Family <- sub("^f__", "", df$Family)

# Identify top 8 most abundant families across all samples
top_families <- df %>%
  group_by(Family) %>%
  summarise(total_abundance = sum(Abundance)) %>%
  arrange(desc(total_abundance)) %>%
  slice_head(n = 8) %>%
  pull(Family)

# (Optional) Manually specify top families to control ordering/consistency
top_families <- c("Enterobacteriaceae", "Enterococcaceae", "Bacteroidaceae",
                  "Oscillospiraceae", "Lachnospiraceae", "Coriobacteriaceae",
                  "Bifidobacteriaceae", "Ruminococcaceae")

# Define color palette for top families and "Other"
families_colors <- c("#a6cee3","#1f78b4","#b2df8a","#33a02c","#fb9a99",
                  "#e31a1c","#fdbf6f","#ff7f00","grey") 

# Map colors to taxa names
families_colors <- setNames(families_colors, c(top_families, "Other"))

# Create stacked barplot of relative abundance per sample
rel_ab_plot <- ps %>%
  microbiomeutilities::aggregate_top_taxa2(top = 8, level = "Family") %>%
  microbiome::plot_composition(otu.sort = "abundance",           # sort taxa by abundance
                               sample.sort = "Bacteroidaceae",   # order samples by one taxon
                               group_by ="sick") +               # group samples by disease status
  scale_fill_manual(values = families_colors, 
                    name = "Family",
                    breaks = c(top_families, "Other")) +
  theme(panel.background = element_blank(),
        axis.text.x = element_text(size = 6, vjust = 1, hjust = 1, angle = 70,  colour = "black"),
        axis.text.y = element_text(size = 9, colour = "black"),
        axis.title.y = element_text(size = 9, colour = "black", vjust = +1),
        axis.title.x = element_text(size = 9, colour = "black", vjust = -1),
        axis.ticks.length.y = unit(.125, "cm"),
        axis.ticks.x = element_blank()) +
  theme(legend.position = "bottom", legend.box = "horizontal",legend.justification = "centre",
        legend.text = element_text(colour = "black", size = 7)) +
  theme(legend.title = element_text(colour = "black", size = 7, face = "bold")) +
  theme(panel.border = element_blank(), axis.line = element_line(colour = "black")) +
  labs(x = "Participant", y = "Relative abundance (%)") +
  theme(plot.margin = margin(1.2, 1.2, 1.2, 1.2, "cm"))+
  theme(strip.text = element_text(angle = 0, size = 10, face = "bold", hjust = 0.5))+
  theme(strip.background = element_blank())  

rel_ab_plot

# Save the plot
ggsave(rel_ab_plot, file = "sample_rel_ab_plot.png",width = 20, height = 13, units = "cm", dpi = 300)
```


```{r}
# ============================================
# Community composition (group-level summary)
# ============================================
# Aggregate to Family level
ps_family <- ps %>%
  tax_glom(taxrank = "Family")

# Convert phyloseq object to long format
df <- psmelt(ps_family)

# Clean Family names (remove "f__" prefix for readability)
df$Family <- sub("^f__", "", df$Family)

# Keep only top families; group all others as "Other"
df$Family <- ifelse(df$Family %in% top_families, df$Family, "Other")

# Calculate mean relative abundance per group (SM vs CC)
df_summary <- df %>%
  group_by(sick, Family) %>%
  summarise(mean_abundance = mean(Abundance), .groups = "drop")

# Create stacked barplot showing average community composition per group
rel_ab_group_plot <- ggplot(df_summary, aes(x = sick, y = mean_abundance, fill = Family)) +
  geom_bar(stat = "identity", position = "stack") +
  scale_fill_manual(values = families_colors) +
  labs(x = "Group", y = "Relative abundance (%)") +
  theme_classic() +
  theme(
    legend.position = "right",
    legend.text = element_text(size = 8),
    legend.title = element_text(size = 9, face = "bold"))

rel_ab_group_plot

# Save the plot
ggsave(rel_ab_group_plot, file = "group_rel_ab_plot.png",width = 8, height = 9, units = "cm", dpi = 300)

```


```{r}
# ============================================
# Alpha diversity (within-sample diversity)
# ============================================
# Fix taxonomy labels (ensures consistent naming across ranks)
ps_fixed <- ps %>%
  tax_fix()

# Calculate alpha diversity metrics (Shannon = diversity, Observed = richness)
alpha_stat_met <-  ps_fixed %>%
  ps_calc_diversity(rank = "Species", index = 'shannon') %>%
  ps_calc_richness(rank = "Species", index = 'observed') %>%
  samdat_tbl() # merge diversity metrics with sample metadata

# Define group comparison (Severe Malaria vs Community Controls)
my_comparisons <- list(c("SM", "CC"))

# Ensure correct ordering of groups on x-axis
alpha_stat_met$sick <- factor(alpha_stat_met$sick, levels = c("SM", "CC"))

# Plot Shannon diversity by group with statistical (Wilcoxon) comparison
alpha_div_plot <- ggplot(alpha_stat_met, aes(x = sick, y = shannon_Species, fill = sick)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  geom_jitter(width = 0.2, size = 1.5, alpha = 0.8) +
  theme_classic(base_size = 12) +
  labs(x = "Group", y = "Shannon index") +
  scale_fill_manual(name = "Group",values = c("SM" = "#00BFC4","CC" = "#F8766D")) +
  stat_compare_means(comparisons = my_comparisons,
                     method = "wilcox.test",
                     label = "p.format",   # shows actual p-value
                     paired = FALSE)

alpha_div_plot

# Save plot
ggsave("alpha_diversity_plot.png",alpha_div_plot,width = 5,height = 5,dpi = 600)

```


```{r}
# ============================================
# BETA DIVERSITY (between-sample differences)
# ============================================
# Fix taxonomy labels for consistency
ps_fixed <- ps %>%
  tax_fix()

# Filter low-prevalence taxa and aggregate to Species level
ps_species <- ps %>%
  tax_fix() %>%                        # fix missing taxonomy
  tax_filter(min_prevalence = 1/100, use_counts = FALSE, verbose = FALSE) %>%
  tax_agg(rank = "Species") # can use Genus

# Apply CLR transformation (accounts for compositional data structure)
ps_clr <- ps_species %>%
  tax_transform("clr", zero_replace = "halfmin")

# Perform beta diversity analysis:
# - Euclidean distance on CLR data = Aitchison distance
# - PCoA visualises differences between samples
beta_2 <- ps_clr %>%
  dist_calc("euclidean") %>%
  ord_calc(method = "PCoA") %>%
  ord_plot(color = "sick", alpha = 0.6, size = 2) +
  theme_bw() +
  coord_fixed(0.7) +
  scale_color_manual(
    name = "Group",
    values = c("SM" = "#00BFC4", "CC" = "#F8766D"))
beta_2

# PERMANOVA test: 
# assesses whether community composition differs by group (and other covariates)
perm_res <- ps_clr %>% dist_calc("euclidean") %>%
  dist_permanova(variables = c("sick", "site", "age", "stool_day"),
  n_perms = 999, seed = 123) %>%
  perm_get()
perm_res

# Extract PERMANOVA results (p-value and R² ) for 'sick'
pval <- perm_res["sick", "Pr(>F)"]
r2    <- perm_res["sick", "R2"]

# Format annotation text for plot
label_text <- paste0("PERMANOVA\nR² = ", round(r2,3), "\np = ", signif(pval,3))

# Add PERMANOVA results to plot (bottom-right corner)
beta_div_plot <- beta_2 +
  annotate("text",
           x = Inf, y = -Inf,
           label = label_text,
           hjust = 1.1, vjust = -0.5,
           size = 2) +
  labs(caption = NULL)
beta_div_plot

# Save plot
ggsave("beta_diversity_plot.png",beta_div_plot,width = 5,height = 5,dpi = 600)

```


```{r}
# ==========================================================
# Differential abundance and prevalence analysis (MaAsLin3)
# ==========================================================
library(tidyverse)
library(maaslin3)

# Load Sylph relative abundance data and sample metadata
sylph_data <- read.delim("Bednarski_20025_sylph_clade_relative_abundance_matrix.tsv",
                         header = TRUE,check.names = FALSE,stringsAsFactors = FALSE)
study_metadata <- read.csv("Bednarski_2025_cleaned_metadata_final.csv",stringsAsFactors = F, sep=",")

# Extract taxa names and abundance matrix
taxa <- sylph_data$clade_name # taxonomic labels
abund <- sylph_data[, -1] # abundance values (samples as columns)

# Transpose matrix so rows = samples, columns = taxa (required for MaAsLin3)
abund_t <- as.data.frame(t(abund))
colnames(abund_t) <- taxa

# Check for duplicated taxa names (can cause model issues)
sum(duplicated(taxa))
taxa[duplicated(taxa)]

# Identify shared sample IDs between abundance data and metadata
common_ids <- intersect(study_metadata$metagm_id, rownames(abund_t))

# Subset both datasets to matching samples
metadata_sub <- study_metadata[study_metadata$metagm_id %in% common_ids, ]
features <- abund_t[common_ids, ]

# Assign rownames to metadata and ensure matching order
rownames(metadata_sub) <- metadata_sub$metagm_id
metadata_sub <- metadata_sub[common_ids, ]

# Final check: ensure sample order matches between datasets
all(rownames(features) == rownames(metadata_sub))

# Inspect feature names (taxonomic strings)
colnames(features)

# Subset to species-level features (recommended for interpretation)
features_species <- features[, grepl("\\|s__[^|]+$", colnames(features))]
colnames(features_species)

# (Optional) subset to genus level for comparison
# features_genus <- features[, grepl("\\|g__[^|]+$", colnames(features))]

# Fit MaAsLin3 model
# Tests association between microbial features and disease status (sick)
fit_out <- maaslin3(
  input_data = features_species,
  input_metadata = metadata_sub,
  output = "maaslin3_output",
  fixed_effects = c("sick"), # variable of interest (SM vs CC)
  normalization = "CLR", # accounts for compositional data
  transform = "NONE", # required when using CLR
  min_abundance = 0.001, # filter very low-abundance features
  min_prevalence = 0.1, # keep features present in ≥10% samples
  warn_prevalence = FALSE,
  cores = 1)

# Load significant results
# Contains features significantly associated with 'sick' (FDR-adjusted)
signif_res <- read.delim("maaslin3_output/all_results.tsv") %>% 
   filter(qval_individual < 0.25)

```

## Reproducibility

To run this module:

1. Clone the repository
2. Place input files in you working directory
3. Open the `.Rmd` file in RStudio
4. Click **Knit → PDF**

Note: MaAsLin3 outputs are precomputed.
